Grouping and summarizing

summarizing the median life expectancy

```{r}
gapminder %>%
  summarize(medianLifeExp = median(lifeExp))


#Filter for 1957 then summarize life expectancy
gapminder %>%
  filter(year==1957) %>%
  summarize(medianLifeExp = median(lifeExp))

gapminder %>%
  filter(year==1957) %>%
  summarize(medianLifeExp = median(lifeExp), maxGdpPercap = max(gdpPercap))

```

Summarizing by year

```{r}
gapminder %>%
  group_by(year) %>%
  summarize(medianLifeExp = median(lifeExp), maxGdpPercap=max(gdpPercap))



```


sumarizing by continent
```{r}
gapminder %>%
  filter(year==1957) %>%
  group_by(continent) %>%
  summarize(medianLifeExp = median(lifeExp), maxGdpPercap= max(gdpPercap))
  arrange(desc(maxGdpPercap))

```
```{r}
gapminder %>%
  group_by(continent, year) %>%
  summarize(medianLifeExp = median(lifeExp), maxGdpPercap= max(gdpPercap))
```
Visualizing median life expectancy over time

```{r}

by_year <- gapminder %>%
  group_by(year) %>%
  summarize(medianGdpPercap = median(gdpPercap))

by_year

ggplot(by_year, aes(x=year, y=medianGdpPercap)) +
  geom_line()
```
```{r}
by_year_continent <- gapminder %>%
  group_by(year,continent)%>%
  summarize(medianGdpPercap = median(gdpPercap))

by_year_continent


ggplot(by_year_continent, aes(x=year,y=medianGdpPercap, color=continent)) +
  geom_line()
```
Bar Plots

visualizing medina GDP per capita by continent

```{r}
by_continent <- gapminder %>%
  filter(year==1952) %>%
  group_by(continent) %>%
  summarize(medianGdpPercap = median(gdpPercap))

by_continent

ggplot(by_continent, aes(x=continent, y=medianGdpPercap)) +
  geom_col()
```
Visualizing GDP per capita by contry in Oceania

```{r}
oceania_1952 <- gapminder %>%
  filter(continent == "Oceania", year == 1952)

oceania_1952

ggplot(oceania_1952,aes(x=country, y=gdpPercap))+
  geom_col()


```
Histogram

```{r}
library(gapminder)
library(ggplot2)
library(dplyr)

gapminder

gapminder_1952 <- gapminder %>%
  filter(year==1952)%>%
  mutate(pop_by_mil = pop/10000000)

gapminder_1952

ggplot(gapminder_1952, aes(x=pop_by_mil))+
  geom_histogram(bins = 50)


```


Data Manipulation With dplyr
```{r}
library(tidyverse)
counties <- read_csv('counties.csv')
glimpse(counties)

counties

```
Select columns
```{r}

counties %>%
  select(state,county, population,men,women)

counties_selected <- counties %>%
  select(state,county,population,private_work,public_work,self_employed)

counties_selected %>%
  arrange(public_work)

counties_selected <- counties %>%
  select(state,county,population)

counties_selected %>%
  filter(population>1000000)

counties_selected <- counties %>%
  select(state,county,population,private_work,public_work,self_employed)

counties_selected %>%
  filter(state =="Texas", population >10000) %>%
  arrange(desc(private_work))

```
select mutate filter and arrange

```{r}
counties %>%
select(state,county,Population,men,women)
mutate(proportion_men = men)

```

the verb count

```{r}
select_counties <- counties %>%
  select(state, region, county,population,citizens)
  
cpimtoes_selected <- counties%>%
  count(region, sort =TRUE) %>%
  arrange(desc(n))
 

```
Counting citizens by state

```{r}
select_counties <- counties %>%
  select(county,region,state,population,citizens)

counties_selected %>%
  count(state, wt = citizenz) %>%
  arrange(desc(n))

```


mutate and counting

```{r}
select_counties <- counties %>%
  select(county,region,state,population,walk)

counties_selected %>%
  mutate(population_walk = population*walk/100) %>%
  count(state, wt=population_walk, sort =TRUE)
  
```
Sumarizing

```{r}
counties_selected <- counties %>%
  select(county,population,income,unemployment)


counties_selected %>%
  summairze(min_population =min(population), max_unemployment = max(unemployment), averagen_income = mean(income))
```
Summarizing by state

```{r}

counties_selected <- counties %>%
  select(state, county,population,land_area)

counties_selected %>%
  group_by(state) %>%
  summarize(total_area=sum(land_area), total_population = sum(population))%>%
  mutate(density=total_population/total_area)
  arrange(desc(density))

  
  counties_selected %>%
    group_by(region,state) %>%
    summarize(total_pop=sum(population)) %>%
    summarize(average_pop=mean(total_pop), median_pop=median(total_pop))

```
Top_n verb

```{r}
counties_selected <- counties %>%
  select(region,state,county,metro,population,walk)

counties_selected %>%
  group_by(region) %>%
  top_n(1, walk)

counties_selected <- counties %>%
  select(region,state,county,metro,population,income)

counties_selected %>%
  group_by(region,state) %>%
  summarize(average_income=mean(income)) %>%
  top_n(1,average_income)

```

```{r}
counties_selected <- counties %>%
  select(state,metro,population)

counties_selected %>%
  group_by(state,metro) %>%
  summarize(total_pop = sum(population)) %>%
  top_n(1,total_pop) %>%
  count(metro)


```
Selecting columns

```{r}
counties %>%
  select(state,county,population,professional) %>%
  arrange(desc(serivce))

counties %>%
  select(state,county,population,ends_with("work")) %>%
  filter(public_work >= 50)

counties %>%
   select(state, county, population, starts_with("income"))

counties %>%
  select(state,county,population,contains("at"))



```
Renaming a colum after count

```{r}
counties %>%
 count(state) %>%
   rename(num_counties = n)

 
 counties %>%
   select(state, county, poverty_rate = poverty)
 
 
```
Using transmute

```{r}
counties %>% 
  transmute(state, county, population, density=population/land_area)
  filter(population >= 1000000) %>%
  arrange(desc(density))

```
Choosing among the four verbs

```{r}
#Change the name of the unemployment column

counties %>%
  rename(unemployment_rate=unemployment)

# keep the state and county columns, and the columns containing poverty

counties %>%
  select(state,county,contains("poverty"))

# Calculate the franction_women column without dropping the other columns

counties %>%
  mutate(fraction_women = women/population)

# keep only the state, county, and employment_rate columns

counties %>%
  transmute(state, county, employment_rate = employed/population)
```

